Demers Cartogram with Rivers

### 2023-09-28 paper/elsarticle-template-num.tex
feat: submitted

\documentclass[Afour,sagev,times]{sagej}

\newcommand{\new}[1]{\textcolor{blue}{#1}}

\usepackage{moreverb,url}
\usepackage{float,hhline,array,tabulary,tabularx,etoolbox,multirow, multicol}

\newcommand{\algorithmautorefname}{Algorithm}

\usepackage[table, dvipsnames]{xcolor}

% \include{preamble/preamble.main}

@@

\newcommand{\bobgraph}[1]{\noindent\textbf{#1}}

% cite with author names
\newcommand{\citea}[1]{\citeauthor{#1}~\cite{#1}}

@@

      \setlength{\intextsep}{3pt}
      \setlength{\columnsep}{3pt}
      \begin{wrapfigure}{r}{0.4\columnwidth}
        \vspace{-15pt}
        \begin{center}
          \includegraphics[width=0.38\columnwidth]{"#1".png}
        \end{center}
      \end{wrapfigure}
      #2

@@

\usepackage{algpseudocode}
\usepackage{algorithm}
% switch for algorithmic
% New definitions
\algnewcommand\algorithmicswitch{\textbf{switch}}
\algnewcommand\algorithmiccase{\textbf{case}}
\algnewcommand\algorithmicdefault{\textbf{default}}

@@

% list with styles
\usepackage{enumitem}
% \include{preamble/preamble.main}

% \include{preamble/preamble.math}
\usepackage{amsmath,bm}
\usepackage{amssymb}

@@

\newcommand{\vectorAvg}{\bm{v}_{avg}}
\newcommand{\NxR}{\node_{n_{x}r}}
\newcommand{\RxN}{\node_{r_{x}n}}
% \include{preamble/preamble.math}

\usepackage[colorlinks,citecolor=red,urlcolor=red]{hyperref}

@@

\setcounter{secnumdepth}{3}

@@

@@

\begin{abstract}
    Cartograms serve as representations of geographical and abstract data, employing a value-by-area mapping technique.
    As a variant of the Dorling cartogram, the Demers cartogram utilizes squares instead of circles to represent regions.
    This alternative approach allows for a more intuitive comparison of regions, utilizing screen space more efficiently.
    However, a drawback of the Dorling cartogram and its variants lies in the potential displacement of regions from their original positions, ultimately compromising legibility, readability, and accuracy.
    To tackle this limitation, we propose a novel hybrid cartogram layout algorithm that incorporates topological elements, such as rivers, into Demers cartograms.
    The presence of rivers significantly impacts both the layout and visual appearance of the cartograms.
    Through a user study conducted on an Electronic Health Records (EHR) dataset, we evaluate the efficacy of the proposed hybrid layout algorithm.
    The obtained results illustrate that this approach successfully retains key aspects of the original cartogram while enhancing legibility, readability, and overall accuracy.
\end{abstract}

\keywords{Cartogram, Demers Cartogram}

\maketitle

\section{Introduction and Motivation}

\begin{samepage}

    Cartograms are representations of geographical and abstract data based on a value-by-area mapping combining statistical and geographical information \cite{dent2009Cartography,inoue2011New}.
    Various styles of cartograms have been proposed and implemented for applications such as urban planning \cite{harris2018Mapping, arranz-lopez2021Enduser}, natural hazard forecasting \cite{pappenberger2019Cartograms, park2020Flood}, conservation and environmental planning \cite{galluzzi2018Mapping, rocchini2019Cartogramming}, political and social demographics \cite{breitzman2018Using, alieva2021How}, and decision-making for public health \cite{gao2020Visualising, sack2021Visualizing}.

    \begin{figure}[b!]
        \centering
        \includegraphics[width=\linewidth,keepaspectratio]{cover.png}
        \caption{The River Thames passing through a cartographic representation of the NHS CCGs in London and surrounding areas.}
        \label{fig:cover}
    \end{figure}

\end{samepage}

@@

\bobgraph{Definitions: }While we focus on rectangular cartogram variants, we start with brief definitions of contiguous and non-contiguous cartograms: Contiguous cartograms preserve topology, maintain connectivity with their adjacent neighbors, but are also subject to distortion in shape.
Non-contiguous cartograms sacrifice topological connectivity with neighbors to enable expansion or reduction in size while maintaining their polygonal shape \cite{dent2009Cartography}.

@@

\thumbnail{dorling}{
    Dorling cartograms, as a variant of non-contiguous cartograms, generally do not preserve geography and topology.
    A Dorling cartogram is statistically accurate, regions are represented by circles and the data dimensions of interest are represented by the circle area \cite{dorling2011Area}.
    In a Demers cartogram, a variant of Dorling, squares are used instead to capture a certain level of topology, as described by \citet{cano2015Mosaic} in their related work section.
    Dorling cartograms are unable to maintain topological accuracy as circles are often repositioned to remove overlap.
    Here we focus on Demers cartograms as we use squares to depict regions.
    This style of cartogram offers the advantages that the comparisons between regions are intuitive and screen space utilization is more efficient.
    This is important in our use case scenario involving EHRs.
    Demers cartograms, where regions are represented by squares, often have the advantage of preserving a higher level of topology at the cost of geographical accuracy \cite{ian2002Cartogram}.
}

\thumbnail{rectangular}{
    Rectangular cartograms are contiguous and do not preserve geographical accuracy \cite{raisz1934Rectangular, vankreveld2004Rectangular}.
    Depending on the variant, a rectangular cartogram may trade-off between statistical and topological accuracy.
}

\thumbnail{mosaic}{
    Mosaic cartograms usually use square or hexagonal tiles to depict regions, and are contiguous and sacrifice statistical accuracy to preserve some level of geographical accuracy \cite{cano2015Mosaic}.
    Some variants are able to preserve topological accuracy as well.
}

@@

\bobgraph{Peer-reviewed Applications: } There has been a substantial amount of research done in this area, here we review some of the important work that has inspired our work.\citet{warf2008Geography} use a Dorling cartogram to represent religious diversity in the US.
\citet{sun2010Effectiveness} depict 1996 US election data and 2005 Chinese population data using Dorling, Mosaic, and contiguous cartograms.
\citet{gao2020Visualising} present a Dorling cartogram to illustrate COVID-19 infections in China.
\citet{tong2018Cartograms} use a Demers cartogram to visualize health-related data by CCG regions in England, the work introduces a novel technique to remove the overlap of squares based on topological features, aiming to improve both geographical and topological accuracy.
\citet{nusrat2020Recognition} investigate the memorability of contiguous and Dorling cartograms using multiple data sets that include demographic, agriculture, and retail data in the US.
See \Cref{table:region vs node} for a list of literature that adopts cartograms for visualization with corresponding geographical regions and node counts.

@@

\bobgraph{Cartograms in Media: }Cartograms are an engaging visual representation and therefore they are a popular choice of representation in covering various topics by the media.
The Washington Post uses cartograms to visualize the US overseas economic assistance, in arm sales (Mosaic) \cite{bearak2016Everything}, the 2016 US Election (contiguous) \cite{gamio2016Election}, and the Brexit Referendum (Mosaic) \cite{taylor2016What}.
National Geographic uses contiguous and Mosaic cartograms to analyze the 2016 US Election results \cite{miller2016Election}, the same topic is also covered by the Financial Times with a Dorling cartogram \cite{stabe2016Search}.
\citet{cruz2017Adapted} adapts a Dorling cartogram with both contiguous and non-contiguous cartograms to represent the gender pay gap in Portugal.
\citet{sandberg2018Cartogram} reports the 2018 US midterm Election with a Mosaic cartogram, the same approach is used to cover the 2020 US Election by the New York Times \cite{thelearningnetwork2020What} and Bloomberg \cite{mccartney20202020}.

@@

\bobgraph{Choropleth Shapefile:} Clinical Commissioning Groups (CCGs) are the primary administrative and geographic unit of the National Health Service (NHS) in the UK \cite{nhsNHS}.
The number of CCGs changes over time due to NHS reorganization.
The most up-to-date shapefile is available from the Open Geography Portalx \cite{opengeographyportalxOpen}.
We decided to use the CCG shapefile from 2020 at the time of writing, due to the absence of published public EHR data based on the latest CCG reorganizations that took place in 2021 and 2022.

\bobgraph{River Shapefiles:} We used OpenStreetMap \cite{openstreetmapRelation} as our data source to obtain shapefiles for the Thames River, the Trent River, and the Great Ouse River in England.
These rivers were chosen as they are well-known rivers, pass through regions with dense populations, and provide informative geographical and topological cues.
Although including smaller rivers is technically feasible, it may not increase the legibility of the cartogram.
This is an open question for future work.

@@

\bobgraph{EHR Data: }We obtained the Clinical Commissioning Group Outcomes Indicator Set (CCG OIS) from NHS Digital \cite{nhsdigitalClinical}.
The OIS is a set of indicators that are used to measure the quality of care and the associated health outcomes in the NHS.
Some datasets include:
\begin{itemize}
    \setlength\itemsep{0px}
    \item Under 75 mortality: cardiovascular disease, respiratory disease, liver disease, and cancer
    \item Emergency hospital admission: stroke, alcohol-specific admission and readmission, coronary heart disease, re-admissions within 30 days of discharge, children with lower respiratory tract infections
\end{itemize}

@@

\bobgraph{Initialization with Rivers:} We first load and (optionally) render the CCG geospatial boundaries.
For each CCG we compute the centroid and represent it using a square node, $ \node $, with the initial size, $ \nodeSize $ =1 pixel.
We then load the river shapefiles and render the rivers.
Since the vertices of the river in the shapefiles are not in sequential order, we first render the starting vertex, followed by the next nearest vertex.
We do not need the original river resolution to incorporate them into cartograms.
We reduce their resolution to match that of the cartogram nodes in order to facilitate node-river intersection tests.
This rendering approach enables us to adjust the river resolution as shown in \Cref{fig:river resolution}.
We further apply simplification by removing vertices that are too close to each other.
The initialization procedure is a one-time process that can be saved for reuse.

\bobgraph{Node Layout and Overlap Removal:} We first apply the Fast Node Overlap Removal (FNOR) algorithm that solves the Variable Placement with Separation Constraints (VPSC) problem \cite{dwyer2006Fast} in order to remove overlap between square nodes.
We chose FNOR over other node overlap removal algorithms because FNOR is capable of minimizing spread and node movement while maintaining a good level of global shape preservation \cite{chen2020Node}.
Initialized with a pixel size of unity, we gradually increase the node size by one unit at a time to ensure smooth transitions.
An increase in $ \nodeSize $ can cause the nodes to overlap.
During overlap removal, we compute node trajectories (See \Cref{alg:TranslateNode}) and translate nodes to their new position.
Nodes that cross a river, denoted $ \NxR $, are translated back to their previous position.
If a node oscillates across a river, we identify this as a stalemate.
One iteration of the layout ends when 1) no node overlap is present; AND 2) no nodes cross a river.
We then increase $ \nodeSize $ by one unit and repeat the algorithm until the average cartographic error, $ \nodeCartographicError $, a measure set by the user, reaches its maximum value $ \nodeCartographicErrorMax $.
The gradual size increase process provides stability to the layout, as can be seen in the accompanying video.

@@

We conduct a \new{within-subject} user study to evaluate the effectiveness of our approach.
We chose this type of evaluation because legibility is a human-centred characteristic.
Many different types of statistical error metrics have been evaluated in previous work \cite{nusrat2016State}, however, our focus is more user-centric in nature.

@@

\bobgraph{Independent Variables: }The primary independent variable is the presence of rivers in the cartograms, which directly impacts the final layout of the cartogram. 
\new{The identical set of rivers is either rendered or hidden collectively.}

@@

135 CCGs are rendered on the screen as a choropleth on the left.
We then generated another view on the right using cartograms.
The color of both visual designs is mapped to a data dimension.
\new{No cartograms were used more than once per participant.}
See \Cref{fig:task} for an example.

@@

Due to pandemic restrictions, we designed the user study to be carried out online.
Participants are expected to use their \new{retail} hardware for the tasks.
However, devices with small screens are not recommended because of the large size of the cartograms that are rendered.
The \new{within-subject} user study includes four parts:

\textbf{P1:} The participants are asked to listen to instructions and training provided as both text and videos.
The instructions are designed to help participants understand the concepts used in the tasks.
Instructional videos are available at \url{https://www.youtube.com/playlist?list=PLL7sHvxLtD75fMtrUQrAdddjt3wfFkcWz}.

\textbf{P2:} The participants are given three practice tasks to familiarize themselves with the user study design.
A sample task is shown in \Cref{fig:task}, and accessible at \url{https://ghr.wangqiru.com/#/P1} and included in the online materials\footnotemark{}.
\new{The practice tasks are identital to the actual tasks, accompanied by instructions provided before the actual tasks commence.
The results from these practice tasks are not included in the final analysis.}
These tasks also serve as a quality test to see whether users are taking the study seriously.
A demonstration of three sample tasks is also included in the instructional video.

\textbf{P3:} The participants are asked to complete 16 location tasks that involve 4 targetCCGs.
\new{See the online materials for exact locations of target CCGs\footnotemark[\value{footnote}].}
Response and reaction time are recorded.
These 4 CCGs are selected to avoid extreme cases (thus bias the result), in terms of size, color, and location.

\textbf{P4:} The participants are asked to complete a questionnaire\footnotemark[\value{footnote}] that consists of \new{5-Point} Likert Scale questions and open-ended questions.

\footnotetext{The materials for P1-P4 are available on the Open Science Foundation website at \url{https://osf.io/q39w7}.}

@@

\noindent\textbf{Color map choice:} The first limitation is the color map that we use to depict the data in our cartograms.
    We use D3's built-in interpolateRdYlGn color map, a diverging color scheme of red, yellow, and green.
    However, we believe that the choice of color map can have a significant impact on the legibility of cartograms.
    In the user study, we carefully avoid extreme values where the location or color of the CCG makes it easier to locate the target.
    See \Cref{fig:extreme} for an example. We plan to explore the impact of different color maps on the legibility of cartograms in future work.

@@

\noindent\textbf{Overlap removal algorithm choice:} Another limitation is the algorithm (FNOR) we use.
We believe that developing a new algorithm with built-in constraint support can significantly reduce the time required to generate cartograms with rivers.
Currently, the runtime of our layout algorithm is approximately 30 milliseconds for each iteration. When the quantity of nodes and features increases, generating the optimal layout demands several hundred iterations.

\noindent\textbf{Generalizability:} Future work also includes generalizations and extensions of the algorithm, e.g., the use of other features in the cartogram layout such as additional rivers, major highways, lakes, and coastlines, etc.
We also consider whether increasing the length of the rivers as the size of the nodes increases would be a useful option.
We would like to explore the case of river-river intersections (or confluence) and testing out more geographies such as the U.S.
and Europe, which have more complex rivers.
We also considered the idea of deforming the rivers as part of the layout algorithm, however, this idea is open to future work.

\noindent\textbf{Improved User Study:} Due to pandemic restrictions, the user study did not take place in a controlled environment.
In the future, we aim to execute a more controlled laboratory study, ensuring a controlled environment and standardized hardware for participants.

@@

\urlstyle{same}
\bibliographystyle{plainnat}
\bibliography{references}

\cleardoublepage

\section{Appendix}
\appendix

\section{Obtaining Shapefile}
\label{obtaining-shapefile}
\begin{lstlisting}[caption={The query that downloads the shapefile of River Thames from OpenStreetMap via the Overpass Turbo API.}, label={overpass},captionpos=b]
    relation(2263653);>>;
    // River Great Ouse: 2798097
    // River Trent: 2863468
    out skel;
\end{lstlisting}
\section{Pre-processing Shapefiles}
\label{app:pre-processing}
Shapefiles from different sources are likely to be incompatible.
In our case, the NHS CCG shapefile is incompatible with the river shapefiles.
The major reason for the incompatibility is the coordinate reference system (CRS).
The CRS of the CCG shapefile is EPSG:27700 (OSGB36 - British National Grid).
The CRS of the river shapefiles is EPSG:4326 (WGS84 - World Geodetic System).
Here, we provide some pre-processing steps using QGIS (version: 3.26.0-Buenos Aires) \cite{qgisWelcome} to handle the incompatibility and reduce shapefile size to improve performance.

@@

\end{document}

>>>>

%% 
%% Copyright 2007-2020 Elsevier Ltd
%% 
%% This file is part of the 'Elsarticle Bundle'.
%% ---------------------------------------------
%% 
%% It may be distributed under the conditions of the LaTeX Project Public
%% License, either version 1.2 of this license or (at your option) any
%% later version.  The latest version of this license is in
%%    http://www.latex-project.org/lppl.txt
%% and version 1.2 or later is part of all distributions of LaTeX
%% version 1999/12/01 or later.
%% 
%% The list of all files belonging to the 'Elsarticle Bundle' is
%% given in the file `manifest.txt'.
%%

%% Template article for Elsevier's document class `elsarticle'
%% with numbered style bibliographic references
%% SP 2008/03/01
%%
%% 
%%
%% $Id: elsarticle-template-num.tex 190 2020-11-23 11:12:32Z rishi $
%%
%%
% \documentclass[preprint,12pt]{elsarticle}

%% Use the option review to obtain double line spacing
% \documentclass[authoryear,preprint,review,12pt]{elsarticle}

%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
%% for a journal layout:
%% \documentclass[final,1p,times]{elsarticle}
\documentclass[5p,times]{elsarticle}
%% \documentclass[final,3p,times]{elsarticle}
%% \documentclass[final,3p,times,twocolumn]{elsarticle}
%% \documentclass[final,5p,times]{elsarticle}
%% \documentclass[final,5p,times,twocolumn]{elsarticle}

%% For including figures, graphicx.sty has been loaded in
%% elsarticle.cls. If you prefer to use the old commands
%% please give \usepackage{epsfig}
\usepackage[table, dvipsnames]{xcolor}

@@

% cite with author names
\newcommand{\citea}[1]{\citeauthor{#1}~\cite{#1}}
\newcommand{\bobgraph}[1]{\noindent\textbf{#1}}

@@

      \setlength{\intextsep}{3pt}
      \setlength{\columnsep}{3pt}
      \begin{wrapfigure}{r}{0.4\columnwidth}
        \vspace{-15pt}
        \begin{center}
          \includegraphics[width=0.35\columnwidth]{"#1".png}
        \end{center}
      \end{wrapfigure}
      #2

@@

% list with styles
\usepackage{enumitem}

%% The amssymb package provides various useful mathematical symbols
\usepackage{amsmath,bm}
\usepackage{amssymb}
%% The amsthm package provides extended theorem environments
%% \usepackage{amsthm}

%% The lineno packages adds line numbers. Start line numbering with
%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
%% for the whole article with \linenumbers.
%% \usepackage{lineno}

\usepackage{float,hhline,array,tabulary,tabularx,etoolbox,multirow,multicol}

\usepackage{moreverb,url}

\newcommand{\algorithmautorefname}{Algorithm}
\usepackage{algpseudocode}
\usepackage{algorithm}
% switch for algorithmic
% New definitions
\algnewcommand\algorithmicswitch{\textbf{switch}}
\algnewcommand\algorithmiccase{\textbf{case}}
\algnewcommand\algorithmicdefault{\textbf{default}}

@@

@@

\newcommand{\vectorAvg}{\bm{v}_{avg}}
\newcommand{\NxR}{\node_{n_{x}r}}
\newcommand{\RxN}{\node_{r_{x}n}}

@@

% \journal{Nuclear Physics B}

@@

\begin{frontmatter}

    %% Title, authors and addresses

    %% use the tnoteref command within \title for footnotes;
    %% use the tnotetext command for theassociated footnote;
    %% use the fnref command within \author or \address for footnotes;
    %% use the fntext command for theassociated footnote;
    %% use the corref command within \author for corresponding author footnotes;
    %% use the cortext command for theassociated footnote;
    %% use the ead command for the email address,
    %% and the form \ead[url] for the home page:

@@

    \begin{abstract}
        Cartograms serve as representations of geographical and abstract data, employing a value-by-area mapping technique.
        As a variant of the Dorling cartogram, the Demers cartogram utilizes squares instead of circles to represent regions.
        This alternative approach allows for a more intuitive comparison of regions, utilizing screen space more efficiently.
        However, a drawback of the Dorling cartogram and its variants lies in the potential displacement of regions from their original positions, ultimately compromising legibility, readability, and accuracy.
        To tackle this limitation, we propose a novel hybrid cartogram layout algorithm that incorporates topological elements, such as rivers, into Demers cartograms.
        The presence of rivers significantly impacts both the layout and visual appearance of the cartograms.
        Through a user study conducted on an Electronic Health Records (EHR) dataset, we evaluate the efficacy of the proposed hybrid layout algorithm.
        The obtained results illustrate that this approach successfully retains key aspects of the original cartogram while enhancing legibility, readability, and overall accuracy.

    \end{abstract}

    \begin{keyword}
        Geospatial visualization \sep Cartogram \sep Demers Cartogram

    \end{keyword}

\end{frontmatter}

%% \linenumbers

%% main text
\section{Introduction and Motivation}
\label{}
Cartograms are representations of geographical and abstract data based on a value-by-area mapping combining statistical and geographical information \cite{dent2009Cartography,inoue2011New}.
Various styles of cartograms have been proposed and implemented for applications such as urban planning \cite{harris2018Mapping, arranz-lopez2021Enduser}, natural hazard forecasting \cite{pappenberger2019Cartograms, park2020Flood}, conservation and environmental planning \cite{galluzzi2018Mapping, rocchini2019Cartogramming}, political and social demographics \cite{breitzman2018Using, alieva2021How}, and decision-making for public health \cite{gao2020Visualising, sack2021Visualizing}.

\begin{figure}[htpb]
    \centering
    \includegraphics[width=0.8\linewidth,keepaspectratio]{cover.png}
    \caption{The River Thames passing through a cartographic representation of the NHS CCGs in London and surrounding areas.}
    \label{fig:cover}
\end{figure}

@@

\subsection{Definitions}

While we focus on rectangular cartogram variants, we start with brief definitions of contiguous and non-contiguous cartograms: Contiguous cartograms preserve topology, maintain connectivity with their adjacent neighbors, but are also subject to distortion in shape.
Non-contiguous cartograms sacrifice topological connectivity with neighbors to enable expansion or reduction in size while maintaining their polygonal shape \cite{dent2009Cartography}.

@@

\thumbnail{dorling}{
    \bobgraph{Dorling cartograms}, as a variant of non-contiguous cartograms, generally do not preserve geography and topology.
    A Dorling cartogram is statistically accurate, regions are represented by circles and the data dimensions of interest are represented by the circle area \cite{dorling2011Area}.
    In a Demers cartogram, a variant of Dorling, squares are used instead to capture a certain level of topology, as described by \citet{cano2015Mosaic} in their related work section.
    Dorling cartograms are unable to maintain topological accuracy as circles are often repositioned to remove overlap.
    Here we focus on Demers cartograms as we use squares to depict regions.
    This style of cartogram offers the advantages that the comparisons between regions are intuitive and screen space utilization is more efficient.
    This is important in our use case scenario involving EHRs.
    Demers cartograms, where regions are represented by squares, often have the advantage of preserving a higher level of topology at the cost of geographical accuracy \cite{ian2002Cartogram}.
}

\thumbnail{rectangular}{
    \bobgraph{Rectangular cartograms} are contiguous and do not preserve geographical accuracy \cite{raisz1934Rectangular, vankreveld2004Rectangular}.
    Depending on the variant, a rectangular cartogram may trade-off between statistical and topological accuracy.
}

\thumbnail{mosaic}{
    \bobgraph{Mosaic cartograms} usually use square or hexagonal tiles to depict regions, and are contiguous and sacrifice statistical accuracy to preserve some level of geographical accuracy \cite{cano2015Mosaic}.
    Some variants are able to preserve topological accuracy as well.
}

@@

\subsection{Peer-reviewed Applications}

There has been a substantial amount of research done in this area, here we review some of the important work that has inspired our work.\citet{warf2008Geography} use a Dorling cartogram to represent religious diversity in the US.
\citet{sun2010Effectiveness} depict 1996 US election data and 2005 Chinese population data using Dorling, Mosaic, and contiguous cartograms.
\citet{gao2020Visualising} present a Dorling cartogram to illustrate COVID-19 infections in China.
\citet{tong2018Cartograms} use a Demers cartogram to visualize health-related data by CCG regions in England, the work introduces a novel technique to remove the overlap of squares based on topological features, aiming to improve both geographical and topological accuracy.
\citet{nusrat2020Recognition} investigate the memorability of contiguous and Dorling cartograms using multiple data sets that include demographic, agriculture, and retail data in the US.
See \Cref{table:region vs node} for a list of literature that adopts cartograms for visualization with corresponding geographical regions and node counts.

@@

\subsection{Cartograms in Media}

Cartograms are an engaging visual representation and therefore they are a popular choice of representation in covering various topics by the media.
The Washington Post uses cartograms to visualize the US overseas economic assistance, in arm sales (Mosaic) \cite{bearak2016Everything}, the 2016 US Election (contiguous) \cite{gamio2016Election}, and the Brexit Referendum (Mosaic) \cite{taylor2016What}.
National Geographic uses contiguous and Mosaic cartograms to analyze the 2016 US Election results \cite{miller2016Election}, the same topic is also covered by the Financial Times with a Dorling cartogram \cite{stabe2016Search}.
\citet{cruz2017Adapted} adapts a Dorling cartogram with both contiguous and non-contiguous cartograms to represent the gender pay gap in Portugal.
\citet{sandberg2018Cartogram} reports the 2018 US midterm Election with a Mosaic cartogram, the same approach is used to cover the 2020 US Election by the New York Times \cite{thelearningnetwork2020What} and Bloomberg \cite{mccartney20202020}.

@@

\subsection{Choropleth Shapefile}

Clinical Commissioning Groups (CCGs) are the primary administrative and geographic unit of the National Health Service (NHS) in the UK \cite{nhsNHS}.
The number of CCGs changes over time due to NHS reorganization.
The most up-to-date shapefile is available from the Open Geography Portalx \cite{opengeographyportalxOpen}.
We decided to use the CCG shapefile from 2020 at the time of writing, due to the absence of published public EHR data based on the latest CCG reorganizations that took place in 2021 and 2022.

\subsection{River Shapefiles}

We used OpenStreetMap \cite{openstreetmapRelation} as our data source to obtain shapefiles for the Thames River, the Trent River, and the Great Ouse River in England.
These rivers were chosen as they are well-known rivers, pass through regions with dense populations, and provide informative geographical and topological cues.
Although including smaller rivers is technically feasible, it may not increase the legibility of the cartogram.
This is an open question for future work.

@@

\subsection{EHR Data}

We obtained the Clinical Commissioning Group Outcomes Indicator Set (CCG OIS) from NHS Digital \cite{nhsdigitalClinical}.
The OIS is a set of indicators that are used to measure the quality of care and the associated health outcomes in the NHS.
Some datasets include:
\begin{itemize}
    \setlength\itemsep{0px}
    \item Under 75 mortality: cardiovascular disease, respiratory disease, liver disease, and cancer
    \item Emergency hospital admission: stroke, alcohol-specific admission and readmission, coronary heart disease, re-admissions within 30 days of discharge, children with lower respiratory tract infections
\end{itemize}

@@

\subsection{Initialization with Rivers}

We first load and (optionally) render the CCG geospatial boundaries.
For each CCG we compute the centroid and represent it using a square node, $ \node $, with the initial size, $ \nodeSize $ =1 pixel.
We then load the river shapefiles and render the rivers.
Since the vertices of the river in the shapefiles are not in sequential order, we first render the starting vertex, followed by the next nearest vertex.
We do not need the original river resolution to incorporate them into cartograms.
We reduce their resolution to match that of the cartogram nodes in order to facilitate node-river intersection tests.
This rendering approach enables us to adjust the river resolution as shown in \Cref{fig:river resolution}.
We further apply simplification by removing vertices that are too close to each other.
The initialization procedure is a one-time process that can be saved for reuse.

\subsection{Node Layout and Overlap Removal}

We first apply the Fast Node Overlap Removal (FNOR) algorithm that solves the Variable Placement with Separation Constraints (VPSC) problem \cite{dwyer2006Fast} in order to remove overlap between square nodes.
We chose FNOR over other node overlap removal algorithms because FNOR is capable of minimizing spread and node movement while maintaining a good level of global shape preservation \cite{chen2020Node}.
Initialized with a pixel size of unity, we gradually increase the node size by one unit at a time to ensure smooth transitions.
An increase in $ \nodeSize $ can cause the nodes to overlap.
During overlap removal, we compute node trajectories (See \Cref{alg:TranslateNode}) and translate nodes to their new position.
Nodes that cross a river, denoted $ \NxR $, are translated back to their previous position.
If a node oscillates across a river, we identify this as a stalemate.
One iteration of the layout ends when 1) no node overlap is present; AND 2) no nodes cross a river.
We then increase $ \nodeSize $ by one unit and repeat the algorithm until the average cartographic error, $ \nodeCartographicError $, a measure set by the user, reaches its maximum value $ \nodeCartographicErrorMax $.
The gradual size increase process provides stability to the layout, as can be seen in the accompanying video.

@@

We conduct a within-subject user study to evaluate the effectiveness of our approach.
We chose this type of evaluation because legibility is a human-centred characteristic.
Many different types of statistical error metrics have been evaluated in previous work \cite{nusrat2016State}, however, our focus is more user-centric in nature.

@@

\bobgraph{Independent Variables: }The primary independent variable is the presence of rivers in the cartograms, which directly impacts the final layout of the cartogram. 
The identical set of rivers is either rendered or hidden collectively.

@@

135 CCGs are rendered on the screen as a choropleth on the left.
We then generated another view on the right using cartograms.
The color of both visual designs is mapped to a data dimension.
No cartograms were used more than once per participant.
See \Cref{fig:task} for an example.

@@

Due to pandemic restrictions, we designed the user study to be carried out online.
Participants are expected to use their retail hardware for the tasks.
However, devices with small screens are not recommended because of the large size of the cartograms that are rendered.
The within-subject user study includes four parts:

\textbf{P1:} The participants are asked to listen to instructions and training provided as both text and videos.
The instructions are designed to help participants understand the concepts used in the tasks.
An instructional video was provided \footnote[1]{Instructional video: \url{https://www.youtube.com/playlist?list=PLL7sHvxLtD75fMtrUQrAdddjt3wfFkcWz}}.

\textbf{P2:} The participants are given three practice tasks to familiarize themselves with the user study design.
A sample task \footnote[2]{A sample task is accessible at \url{https://ghr.wangqiru.com/#/P1}.} is shown in \Cref{fig:task} and included in the online materials\footnotemark[3].
The practice tasks are identital to the actual tasks, accompanied by instructions provided before the actual tasks commence.
The results from these practice tasks are not included in the final analysis.
These tasks also serve as a quality test to see whether users are taking the study seriously.
A demonstration of three sample tasks is also included in the instructional video\footnotemark[1].

\textbf{P3:} The participants are asked to complete 16 location tasks that involve 4 target CCGs.
See the online materials for exact locations of target CCGs\footnotemark[3].
Response and reaction time are recorded.
These 4 CCGs are selected to avoid extreme cases (thus bias the result), in terms of size, color, and location.

\textbf{P4:} The participants are asked to complete a questionnaire\footnote[3]{The materials for P1-P4 are available on the Open Science Foundation website at \url{https://osf.io/q39w7}.} that consists of 5-Point Likert Scale questions and open-ended questions.

@@

\subsection{Color map choice}

The first limitation is the color map that we use to depict the data in our cartograms.
    We use D3's built-in interpolateRdYlGn color map, a diverging color scheme of red, yellow, and green.
    However, we believe that the choice of color map can have a significant impact on the legibility of cartograms.
    In the user study, we carefully avoid extreme values where the location or color of the CCG makes it easier to locate the target.
    See \Cref{fig:extreme} for an example. We plan to explore the impact of different color maps on the legibility of cartograms in future work.

@@

\subsection{Overlap removal algorithm choice}

Another limitation is the algorithm (FNOR) we use.
We believe that developing a new algorithm with built-in constraint support can significantly reduce the time required to generate cartograms with rivers.
Currently, the runtime of our layout algorithm is approximately 30 milliseconds for each iteration. When the quantity of nodes and features increases, generating the optimal layout demands several hundred iterations.

\subsection{Generalizability}

Future work also includes generalizations and extensions of the algorithm, e.g., the use of other features in the cartogram layout such as additional rivers, major highways, lakes, and coastlines, etc.
We also consider whether increasing the length of the rivers as the size of the nodes increases would be a useful option.
We would like to explore the case of river-river intersections (or confluence) and testing out more geographies such as the U.S.
and Europe, which have more complex rivers.
We also considered the idea of deforming the rivers as part of the layout algorithm, however, this idea is open to future work.

\subsection{Improved User Study}

Due to pandemic restrictions, the user study did not take place in a controlled environment.
In the future, we aim to execute a more controlled laboratory study, ensuring a controlled environment and standardized hardware for participants.

@@

\section{Acknowledgements}
This work is funded by the grant EP/S010238/2 from the Engineering and Physical Sciences Research Council (EPSRC). EPSRC is a British Research Council that provides government funding for grants to undertake research and postgraduate degrees in engineering and the physical sciences.

\bibliographystyle{elsarticle-num-names}
\bibliography{references}
\cleardoublepage

\appendix

\section{Obtaining Shapefile}
\label{obtaining-shapefile}
\begin{lstlisting}[caption={The query that downloads the shapefile of River Thames from OpenStreetMap via the Overpass Turbo API.}, label={overpass},captionpos=b]
    relation(2263653);>>;
    // River Great Ouse: 2798097
    // River Trent: 2863468
    out skel;
\end{lstlisting}

\section{Pre-processing Shapefiles}
\label{app:pre-processing}
Shapefiles from different sources are likely to be incompatible.
In our case, the NHS CCG shapefile is incompatible with the river shapefiles.
The major reason for the incompatibility is the coordinate reference system (CRS).
The CRS of the CCG shapefile is EPSG:27700 (OSGB36 - British National Grid).
The CRS of the river shapefiles is EPSG:4326 (WGS84 - World Geodetic System).
Here, we provide some pre-processing steps using QGIS (version: 3.26.0-Buenos Aires) \cite{qgisWelcome} to handle the incompatibility and reduce shapefile size to improve performance.

@@

\end{document}
\endinput
%%
%% End of file `elsarticle-template-num.tex'.
